## More biplots

::: {style="font-size: 120%;"}

-   CVA biplots for two classes
-   Regression biplot
-   Spline biplot
:::

## CVA biplots for two classes
::: {style="font-size: 120%;"}

Canonical space of dimension 1.

Solve $\mathbf{BM=WM\Lambda}$ where $\mathbf{M} = \begin{bmatrix}
                      \mathbf{m}_1 & \mathbf{M}_2\\
             \end{bmatrix}$

$$
\bar{\mathbf{Y}} = \bar{\mathbf{X}} \mathbf{M} = \begin{bmatrix}
                                  \bar{y}_{11} & 0      & \dots & 0 \\
                                  \vdots       & \vdots &       & \vdots\\
                                  \bar{y}_{K1} & 0      & \dots & 0 \\
                                    \end{bmatrix}
$$

$$
\mathbf{\Lambda} = diag(\lambda, 0, ..., 0)
$$ Total squared reconstruction error for means: $TSREM = tr\{ (\bar{\mathbf{X}}-\hat{\bar{\mathbf{X}}})(\bar{\mathbf{X}}-\hat{\bar{\mathbf{X}}})'\} = 0$

Total squared reconstruction error for samples: $TSRES = tr\{ ({\mathbf{X}}-\hat{{\mathbf{X}}})({\mathbf{X}}-\hat{{\mathbf{X}}})'\} >0$

Minimise $TSRES$ (Default option)

Alternative option: Maximise Bhattacharyya distance. For more details see

-   le Roux, N. and Gardner-Lubbe, S., 2024. A two-group canonical variate analysis biplot for an optimal display of both means and cases. *Advances in Data Analysis and Classification*, pp.1-28.

:::

## CVA biplots for two classes

::: {style="font-size: 120%;"}

Solve $\mathbf{BM=WM\Lambda}$ where $\mathbf{M} = \begin{bmatrix}
                      \mathbf{m}_1 & \mathbf{M}_2\\
             \end{bmatrix}$

Minimise $TSRES$

$$
\mathbf{M}^{-1} = \begin{bmatrix}
                      \mathbf{m}^{(1)} \\
                      \mathbf{M}^{(2)}\\
             \end{bmatrix}
$$

$$
\mathbf{M}^{(2)}\mathbf{M}^{(2)'} = \mathbf{UDV}'
$$

$$
\mathbf{M}_{opt} = \begin{bmatrix}
                      \mathbf{m}_1 & \mathbf{M}_2\mathbf{V}\\
             \end{bmatrix}
$$
:::

## CVA biplots for two classes

```{r, CVA2class}
#| fig-height: 7
biplot (iris[51:150,]) |> CVA (classes = iris[51:150,5]) |> means (cex = 2) |>
  axes (label.dir = "Hor", label.line = c(0.8, 0, 0, 0)) |> plot ()
```

## Regression biplot

::: {style="font-size: 120%;"}
Any 2D representation of sample points, for example
:::
```{r}
library (MASS)
Zmat <- sammon (dist(iris[-102,1:4], method = "manhattan"))$points
```

::: {style="font-size: 120%;"}

To create a biplot we need to add information on the variables.

$$
\mathbf{X}:n \times p
$$

$$
\mathbf{Z}:n \times 2
$$

$$
\mathbf{X = ZB + E}
$$

$$
\mathbf{B = (X'X)}^{-1}\mathbf{X'Z}
$$
:::

## Regression biplot

```{r, regression}
#| fig-height: 7
biplot (iris[-102,]) |> regress (Zmat) |>  plot ()
```

## Spline biplot

::: {style="font-size: 120%;"}

Are linear axes a good representation when the transformation from $\mathbf{X}:n \times p$ to $\mathbf{Z}:n \times 2$ is nonlinear?

Replace linear regression with splines.
:::

## Spline biplot

```{r, splines}
#| fig-height: 7
biplot (iris[-102,1:4]) |> regress (Zmat, axes = "splines") |>  plot ()
```

## In conclusion
::: {style="font-size: 120%;"}
Biplots developed since 1971.

R code (SPLUS) developed since 1997.

The **biplotEZ** package encapsulates the developments in a single package.

Developments on **biplotEZ** are continuing

* Categorical variables: Multiple correspondence analysis & Categorical PCA.
* Shiny / plotly version for interactivity.
* Modelling: bi-additive biplots, biplots for logistic regression.

*Applying biplotEZ an R package for multi-dimensional data visualisation*

Contact one of the team members or e-mail muvisu@sun.ac.za.
:::